### 2019-04-16 reports/article.tex
working on the paper

% title goes here:
\begin{flushleft}
{\Large
\textbf\newline{Spatiotemporal patterns of urbanization in three Swiss agglomerations: landscape metrics, growth modes and self-organized criticality}
}
\newline
% authors go here:
\\
Mart\'i Bosch\textsuperscript{1*},
J\'er\^ome Chenal\textsuperscript{1},
St\'ephane Joost\textsuperscript{2,1},
\\
\bigskip
\textbf{1} Urban and Regional Planning Community (CEAT) \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland
\\
\textbf{2} Laboratory of Geographic Information Systems (LASIG) \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland
\\
\bigskip
* Corresponding author: marti.bosch@epfl.ch

@@

\section*{Abstract}
Fitting multiple figures into very tight manuscripts while keeping it pleasant to read is challenging. Therefore figures are often simply attached to the very end of a manuscript file. While easier for the authors, this practice is inconvenient for readers. This \LaTeX template shows how to generate a compiled PDF with figures embedded into the text. It provides several examples of how to embed figures or tables directly into the text thus giving you a range of options from which you should choose the one best suited for your manuscript. Check out Schlegel et al., (2016) as example of use \cite{liu2016general}.

@@

@@

@@

% The evolution of the urban agglomeration
The spatiotemporal evolution of the urban footprint for the three urban agglomerations is displayed in \autoref{figures/landscape_plots.pdf}.

\begin{figure}[!ht]
  \begin{adjustwidth}{-.4\textwidth}{0cm}
    \centering  
    \includegraphics[width=\linewidth]{figures/landscape_plots.pdf}
    \vspace{.5em}
    \caption[Evolution of urban patches]{\label{figures/landscape_plots.pdf}Evolution of urban patches of the three urban agglomerations throughout their respective periods of study. The times t$_0$, t$_1$, t$_2$ and t$_3$ correspond to 1981, 1993, 2004 and 2013 for Bern; 1980, 1990, 2005 and 2014 for Lausanne and 1982, 1994, 2007 and 2016 for Zurich}
  \end{adjustwidth}
\end{figure}

\subsection*{Quantifying spatio-temporal patterns of urbanization}

The landscape metrics selected in this study are described in \autoref{selected-metrics}.

\begin{table}[!h]
  \begin{adjustwidth}{-.4\textwidth}{0cm} % Comment out/remove adjustwidth environment if table fits in text column.
    \footnotesize % Font size can be changed to match table content. Recommend 10 pt.
    \renewcommand{\arraystretch}{1.3} % horizontal spacing
    \centering
    \caption[Selected landscape metrics]{\label{selected-metrics}Selected landscape metrics to quantify the spatio-temporal patterns of urbanization}
    \begin{tabular}{p{.34\textwidth} p{.2\textwidth} p{.14\textwidth} p{.6\textwidth}} 
      \toprule
      \textbf{Metric name (abbrev.)} & \textbf{Level} & \textbf{Category} & \textbf{Description} \\
      \midrule
      Percentage of landscape (PLAND) & Class & Area and edge & Percentage of landscape, in terms of area, occupied by patches of a given class \\
      Mean patch size (MPS) & Class and landscape & Area and edge & Mean patch size \\
      Largest patch index (LPI) & Class and landscape & Area and edge & Percentage of the landscape occupied by the largest patch \\
      Patch density (PD) & Class and landscape & Aggregation & The number of patches per area unit \\
      Edge density (ED) & Class and landscape & Area and edge & Sum of the lengths of all edge segments per area unit \\
      Area-weighted mean fractal dimension (AWMFD) & Class and landscape & Shape & Mean patch fractal dimension weighted by relative patch area \\
      Mean euclidean nearest neighbor distance (ENN) & Class and landscape & Aggregation & Mean patch shortest edge-to-edge distance to the nearest neighboring patch of the same or different class \\
      Landscape shape index (LSI) & Class and landscape & Aggregation & Standardized ratio of edge length to area \\
      Contagion (CONTAG) & Landscape & Aggregation & Metric that measures the extent to which patches of the same class are spatially aggregated in the landscape \\
      Shannon's diversity index (SHDI) & Landscape & Diversity & Metric that measures the diversity of patch types determined by both the number of different patch types and the proportional distribution of area among patch classes. \\
      % SqP
      \bottomrule  
    \end{tabular}
  \end{adjustwidth}
\end{table}

\subsection*{Modes of urban growth}

The different growth modes can be detected by means of the Landscape Expansion Index (LEI) \cite{liu2010new}, which is defined as:

\begin{equation}
  \label{eq:lei}
  LEI = \frac{A_0}{A_0 + A_v}
\end{equation}

\subsection*{Fractal order and self-organized criticality}

@@

% Fractal geometry has been extensively employed to describe morphological characteristics of cities and landscapes.
Cities, like many other geographical phenomena, display complex morphological traits which despite their irregular appearance, comply with well-defined order principles that can be characterized quantitatively by means of fractal geometry.
% Two recurrent empirical observations are of particular interest in the context of the study of land use change associated to urbanization \cite{white2015modeling}.
Two main characteristics of fractal structures are of particular interest in the context of the study of urbanization and land use change \cite{white2015modeling}.

\paragraph*{Area-radius scaling}
% On the one hand, the radial dimension, which describes the relationship between the built-up area of an urban agglomeration and the distance from the main city center, has been shown to empirically conform to the relationship:
On the one hand, the relationship between the built-up area of an urban agglomeration and the distance from the main city center has been shown to empirically conform to the relationship:

@@

where $A$ denotes the total area of the urban built-up extent that lays within a distance $r$ from the city center, and $D$ corresponds to the radial dimension, analogous to the fractal dimension of two-dimensional complex objects such as Sierpinski carpets.
% Although the measure might not be appropriate for urban agglomerations with multiple important centers, \cite{frankhauser1994fractalite} found extensive evidence that not only most contemporary cities could be approximated through (\ref{eq:radial-dimension}), but also that with very few exceptions the value of $D$ fell between 1.9 and 2.0.
Although the measure might not be appropriate for urban agglomerations with multiple important centers, \cite{frankhauser1994fractalite} found extensive evidence that most contemporary cities could be approximated through \autoref{eq:radial-dimension}, with the value of $D$ consistently falling between 1.9 and 2.0.
% Additionally, as noted first by \cite{frankhauser1990aspects} and confirmed in subsequent examination by \cite{white1993cellular},
Additionally, following initial observations by \cite{frankhauser1990aspects}, \cite{white1993cellular} suggested that the area-radious scaling could be better approximated throguh two values of $D$, a first steeper one for small values of $r$, reflecting an inner zone where urbanization was essentially complete, and a second lower slope for the outer zone that is still undergoing urbanization.

@@

\paragraph*{Patch size-frequency distribution}
% On the other hand, the size distribution of urban patches
On the other hand, contemporary urban agglomerations are configured by numerous patches of urban land use.
% For fractal objects, there must be no characteristic patch size
% A hallmark of fractal objects is the self-similarity accros scales, which means that there must be no characteristic patch size.
If such configuration is fractal, there must be no characteristic patch size, and thus the relationship between the size of an urban patch and the number of patches of that size found in the agglomeration must follow a power-law scaling relationship of the form:

\begin{equation}
  \label{eq:radial-dimension}
  N(s) \sim s^{-D}
\end{equation}

where $N(s)$ is the number of patches of size $s$, and the scaling exponent $D$ is the patch size-frequency dimension.

@@

The relative dominance of infilling, edge expansion and leapfrog development during the period of study is displayed in \autoref{figures/growth_modes.pdf}.
% TODO: mention LEI and AWLEI here and in the figure caption
Overall, considering the influence of growth modes in terms of the number of new urban patches rather than their respective area tends to over-represent the impact of leapfrog, since the new urban patches that appear non-adjacent to existing urban patches tend to be small.
% As suggested by \cite{li2013quantifying}, edge-expansion maintains its importance throughout the three periods considered for each agglomeration, whereas leapfrog and infilling alternate their relative dominance.
As suggested by \cite{li2013quantifying}, edge-expansion maintains its importance throughout all agglomerations and time periods, whereas leapfrog and infilling alternate their relative dominance.
% Nevertheless, the way in which the latter happens does not seem random in the present study but rather suggests a tendence towards increasing infill, more noticeable when considering its influence in terms of area.
Nevertheless, in the present study, such alternation does not seem to occur randomly but rather suggests a tendence towards increasing infill, more noticeable when considering its influence in terms of area.
% A potential explanation might stem from the fact that suitable urban land is becoming incresingly scarce.
% See \nameref{code-sensitivity}.

>>>>

% title goes here:
\begin{flushleft}
{\Large
\textbf\newline{Spatiotemporal patterns of urbanization in three Swiss urban agglomerations: insights from landscape metrics, growth modes and fractal analysis}
}
\newline
% authors go here:
\\
Mart\'i Bosch\textsuperscript{1*},
J\'er\^ome Chenal\textsuperscript{1},
St\'ephane Joost\textsuperscript{2,1},
\\
\bigskip
\textbf{1} Urban and Regional Planning Community (CEAT) \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland
\\
\textbf{2} Laboratory of Geographic Information Systems (LASIG) \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland
\\
\bigskip
* Corresponding author: marti.bosch@epfl.ch

@@

\section*{Abstract}

Urbanization is currently a global phenomenon that has become the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions.
In order to evaluate the impacts of urbanization and inform urban planning, it is important to understand the spatiotemporal patterns of land use change associated to urbanization.
This paper exploits three different frameworks, namely landscape metrics, urban growth modes and fractal analysis to characterize the spatiotemporal patterns of urbanization of the Swiss urban agglomerations of Zurich, Bern and Lausanne. % from 1980 to 2016
The land use inventory provided by the Swiss Federal Statistical Office were used to assemble four temporal snapshots from 1980 to 2016 at the extent of the urban agglomerations.
The time series of landscape metrics generally supports the diffusion and coalescence hypothesis, yet many remaining natural patches hinder the coalescence of urban patches.
On the other hand, an analysis of the urban growth modes suggest that leapfrog development occurs at all periods, which contributes to an increasing fragmentation of natural patches and maintains the fractal configuration of the landscape.
The discussion reviews potential explanations for the observed landscape changes, and concludes with some planning implications.

\textbf{Keywords:} urbanization; land use change; spatial pattern analysis; landscape metrics; diffusion and coalescence hypothesis; urban growth modes; fractals; scaling; complexity

@@

% Urbanization has become a global phenomenon
The last centuries have seen an unprecedented growth of urban areas, which has resulted in dramatic conversion of natural land and profound changes in landscape patterns and the ecosystem functions that they support \cite{alberti2005effects}. % fragments, isolates and degrades natural habitats; simplifies and homogenizes species composition; disrupts hydrological systems; and modifies energy flow and nutrient cycling.
% With more than half of the world's population currently living in urban areas,
The combination of current demographic prospects and the observed trends of decreasing urban densities suggest that the global amount of land occupied by cities might increase threefold \cite{angel2005dynamics}.
Quantifying urban landscape patterns in space and time is an important and necessary step to understand the driving forces and ecological impacts of urbanization \cite{wu2014urban}.

Recent decades have witnessed an increasing number of studies of the spatiotemporal patterns of land use change associated to urbanization \cite{dietzel2005spatio, seto2005quantifying, schneider2008compact, jenerette2010global, wu2011quantifying, li2013quantifying, liu2016general, nong2018quantifying}.
% hypotheses:
%% diffusion and coalescence, analogous to multiple nuclei development and envelopment
%% challenged such hypotheses  and concluded
% Initial comparative studies
Initial attempts to synthesis suggested that urbanization can be characterized as a two-step alternating process of diffusion and coalescence \cite{dietzel2005spatio, schneider2008compact}, nonetheless, subsequent studies challanged the empirical validity of such hypothesis.
% sample of 120 global cities from 1990 to 2000.
On the one hand, Jenerette and Potere \cite{jenerette2010global} examined the spatiotemporal patterns of land use change of a sample of 120 global cities from 1990 to 2000, and determined that overall, urbanization leads to fragmented landscapes with more complex and heterogeneous structures. On the other hand, Li et al. \cite{li2013quantifying} determined that the two-phase diffusion and coalescence model can be over-simplistic and that urbanization might be better characterized by means of three growth modes, namely infilling, edge expansion and leapfrogging, which operate simoultaneously while alternating their relative dominance.

Paralleling the above studies, approaches from the complexity sciences have provided novel insights into the spatial organization that underpins contemporary cities \cite{batty2005cities}.
% Although the scaling relationships of fractal geometry provide valuable measures of the extent to which urban patches fill a given landscape \cite{batty1989urban}
% As a matter of fact, urban forms encountered over a wide variety of cities exhibit traits that are consistent with fractal geometry and suggest the existence of strong morphological regularities \cite{frankhauser1994fractalite, batty1994fractal, white2015modeling}.
% Like many other complex systems, cities exhibit morphological traits that are consistent with fractal geometry and might be viewed as evidence of the self-organizing processes that occur upon them \cite{white1993cellular}.
% Although the scaling relationships of fractal geometry suggest the existence of strong morphological regularities accross a wide variety of cities and regions \cite{frankhauser1994fractalite, batty1994fractal, white2015modeling}, they have not been exploited in
Like many other complex systems, cities exhibit morphological traits that are consistent with fractal geometry and reflect the self-organizing nature of the processes that occur upon them \cite{white1993cellular}.
Although the scaling relationships of fractal geometry suggest the existence of strong morphological regularities over a wide variety of cities and regions \cite{frankhauser1994fractalite, batty1994fractal, white2015modeling}, % they have rarely been exploited within the study of the spatiotemporal patterns of urbanization.
their meaning in the context of the spatiotemporal patterns of urbanization remains unclear \cite{li2000fractal, manson2006complexity, bosch2019addressing}.

% TODO: self-organized criticality
% Nevertheless,

This paper intends to evaluate the spatiotemporal patterns of land use change observed in three of the largest Swiss urban agglomerations from three different perspectives, namely landscape metrics, urban growth modes and fractal analysis. 
The objective of this paper is twofold.
The first is to assess to test whether the spatiotemporal evolution of the three agglomerations conform to the diffusion and coalescence dichotomy, the second is to explore how the three adopted perspectives might complement each other.
The results will serve to discuss planning implications regarding the desirability of the recent densification policies adopted in Switzerland.
% The second is to explore how the such  might be related to characteristics of fractal objects.

@@

Switzerland is a highly developed country in central Europe %, geographically divided in three ecoregions, namely the Central Plateau, the Alps and the Jura region. Mainly because of their topography, most of the population is concentrated in its Central Plateau region, which accounts for about one third of the total Swiss territory (42,000 km$^2$).
% with a distribution of population characterized
% whose distribution of population configures a
with a population distributed into several interconnected mid-sized cities and a large number of small municipalities.
% Such polycentric urban system is characterized by a high level of accessibility and short commuting distances.
Mainly because of the country's topography, most urban settlements are located in its Central Plateau region, which accounts for about one third of the total Swiss territory, % (42,000 km$^2$) and is highly urbanized (450 inhabitants per km$^2$).
% Central plateau: elevation ranging from 400 to 700m, situated within a transition zone between humid oceanic climate and continental temperate climate with mean annual temperatures of 9-10 C, with the dominating vegetation being mixed broadleaf forest.
% with elevations ranging from 400 to 700m, a continental temperate climate with mean annual temperatures of 9-10 $^{\circ}$C and mean annual precipitation of 800-1400 mm, and a dominating vegetation of mixed broadleaf forest.
The Central Plateau is characterized by elevations that range from 400 to 700m, a continental temperate climate with mean annual temperatures of 9-10 $^{\circ}$C and mean annual precipitation of 800-1400 mm, and a dominating vegetation of mixed broadleaf forest.

% The Federal Statute on regional planning of 1979 already included the responsibility to avoid sprawl by ensuring that land is used economically and that the extension of settlements must be limited ; also strengthened the role of the designated building zones and clearly reduced the number of new buildings constructed outside them.
% However, not effective because the municipalities can designate new building zones almost entirely autonomously
% Countryside initiative (August 2008) to improve existing regional planning law in two respects: (i) federal gov must promote high-quality internal urban development, and (ii) freeze total designated building zones for 20 years.
% Result: revision of the Federal Statute with (i) designation of building zones limited on anticipated need based on population growth forecast of the next 15 years, and (ii) levies to compensate for increase of property values. Accepted by Swiss voters in March 2013 and countryside withdrawn.
% Distribution of power:
%% federal: framework legislation and coordination
%% cantons: spatial planning (enact laws and comprehensive plans)
%% municipalites: land use plan and building ordinance ; compulsory since the Federal Statute of 1979
In line with the country's federalist government structure, the Swiss spatial planning system is distributed between the federal state, the 26 cantons and 2495 municipalities. The federal state specifies the framework legislation and coordinates the spatial planning activities of the cantons, while cantons check the compliance of municipal development plans with cantonal and federal laws. Municipal administrations are in charge of their local development plans, namely the land use plan and building ordinance, and might therefore be viewed as the most important spatial planning entities.
% Important aspects of spatial planning that are defined within federal laws include the delimitation of the number of new buildings constructed outside the building zone or the definition of quotas of arable land to be maintained for each canton.
While the Federal Statue on regional planning of 1979 limited the number of new buildings constructed outside the building zones, built-up areas have since increased continuously, mainly because the municipalities can designate new building zones almost entirely autonomously \cite{jaeger2014improving}. A major revision of the Federal Statue was accepted in 2013, which limits the municipal competences to designate new building zones and encourages infill development and densification by means of tax incentives.
% Current urbanization trends suggest significant increases of urban land use demands over the next decades, mostly at the expense of agricultural land located at the fringe of existing urban agglomerations \cite{price2015future}.
Forecasts based on the current urbanization trends predict significant increases of urban land use demands over the next decades, mostly at the expense of agricultural land located at the fringe of existing urban agglomerations \cite{price2015future}.

Given that a significant part of the cross-border urban agglomerations of Basel and Geneva lie beyond the Swiss boundaries \cite{sfso2014espace}, in order to ensure coherence of the land use/land cover data (see the section below), this study comprises only three of the five largest Swiss urban agglomerations, namely Zurich, Bern and Lausanne \cite{sfso2018city}.
As shown in \autoref{figures/population_change.pdf}, the three cities have undergone important population growth over the last 30 years, especially during the most recent years and at the agglomeration extent.
With a total population over 1.3 million and land area of 1305 km$^2$ (1038 hab/km$^2$), Zurich is the largest Swiss urban agglomeration. As a leading global city and one of the world's largest financial centers, Zurich has the country's largest airport and railway station, and also hosts the largest Swiss universities and higher education institutions. 
% ranking among the most liveable cities in the world.
Bern is the capital of Switzerland and fourth most populous urban agglomeration in Switzerland, with a total population of 410000 inhabitants and occupying a land area of 783 km$^2$ (531 hab/km$^2$). As the fifth largest Swiss urban agglomeration and the second most important student and research center after Zurich, Lausanne has a total population of 409000 inhabitants over a land area of 773 km$^2$ (537 hab/km$^2$). Given its larger population growth rate, Lausanne is likely to soon surpass Bern and become the fourth largest urban agglomeration in Switzerland.
% Zurich: 1021859 (1990), 1080728 (2000), 1191058 (2010), 1354140 (2017)
% Bern: 351084 (1990), 349157 (1990), 357668 (2010), 415785 (2017)
% Lausanne: 300280 (1990), 311441 (2000), 339,389 (2010), 415596 (2017)
% Densities: Zurich: 1037 Bern: 531, Lausanne: 537
Overall, the three urban agglomerations are characterized by a pervasive public transportation system and a highly developed economy, with a 85\% of the employment devoted to the tertiary sector.

\begin{figure}[!ht]
    \centering  
    \includegraphics[width=\linewidth]{figures/population_change.pdf}
    \vspace{.5em}
    \caption[Population change]{\label{figures/population_change.pdf}Population change of the three regions of study at the city core (left) and agglomeration extent (right) over the periods of 1990-2000, 2000-2010 and 2010-2017. Data from the Urban Audit collection \cite{sfso2018city}.}
\end{figure}

@@

The two main land use/land cover datasets available for Switzerland are the CORINE land cover (CLC) inventory \cite{heymann1994corine} and the Swiss land statistics (SLS) datasets \cite{sfso2017statistique}.
As part of a pan-European project, the CLC provides an inventory of land cover maps which are standardized among a set of European countries.
The CLC inventory comprises a hierarchical nomenclature of 3 levels, where level 3 comprises 44 land cover classes, and datasets have been released for 1990, 2000, 2006, 2012 and 2018, with a spatial resolution of one hectare per pixel. % \footnote{Although Switzerland was not part of the first CLC inventory for the year 1990, Switzerland published in 1998 a compatible dataset for CLC 1990 \cite{steinmeier2013corine}. By the release of the following CLC dataset in 2000, Switzerland was already a member of the CORINE project, therefore the CLC datasets for 2000, 2006, 2012 and 2018 already include the country.}.
% The SLS land use land cover maps are assembled by the Swiss Federal Statistical Office (SFSO) as part of the production process of the national maps 
On the other hand, the SLS inventory provides maps for the national extent of Switzerland which comprise 72 base categories, and four datasets have been released for 1979/85, 1992/97, 2004/09 and 2013/18\footnote{\label{fn:sls-years}The exact dates of each surveying period 1979/85, 1992/97, 2004/09 and 2013/18 are determined according to the production process of the national maps and vary accross the Swiss territory \cite{sfso2017statistique}}, also with the spatial resolution of one hectare per pixel.

Given their pan-European extent, the CLC inventory would allow the examination of the cross-border urban agglomerations of Geneva and Basel, nevertheless, two main reasons favor the use of the SLS datasets in the context of the present study.
On the one hand, Switzerland was not part of the first CLC inventory for 1990, and although a dataset for the Swiss extent compatible with CLC 1990 was released in 1998 \cite{steinmeier2013corine}, it is only available at a coarser resolution of 250m$\times$250m per pixel. Accordingly, employing CLC at the hectare resolution for Switzerland would be possible over the 2000-2018 period. Instead, SLS provides consistent datasets from 1979 to 2018, therefore allowing a more thorough study of the spatiotemporal patterns of urbanization.
% principle of predominant use, which tends to systematically underpredict the human impacts by completely eliminating human features that occupy modest parts
% interpretation by sampling
On the other hand, while both CLC and SLS rely on computer-aided interpretation of satellite imagery in order to produce the land cover maps, CLC is based on the principle of predominant use, which has been reported to systematically underpredict the human impacts by completely eliminating human features that occupy important yet non-dominant portions of the hectometrical pixel \cite{sfso2017statistique}. In response to such shortcomings, SLS adopted a methodology of interpretation by sampling, which includes special treatment and field verification of pixels where the category attribution is not clear. Since this study intends to explore land use changes related to urban sprawl and fragmentation at the urban fringe, the presence of significant human settlements outside the main urban clusters might not be appreciable within the CLC inventory (see \autoref{figures/sls_clc_comparison.pdf}).

\begin{figure}[!ht]
  \begin{adjustwidth}{-.4\textwidth}{0cm}
    \centering  
    \includegraphics[width=\linewidth]{figures/sls_clc_comparison.pdf}
    \vspace{.5em}
    \caption[Comparison of SLS and CLC]{\label{figures/sls_clc_comparison.pdf}Comparison of the SLS and CLC datasets to the register of buildings and dwellings provided by the Swiss Federal Statistical Office (SFSO). The three maps correspond to the same spatial extent, with the city of Bern at its center.}
  \end{adjustwidth}
\end{figure}

% The preparation of the extracts for the urban agglomerations of Zurich, Bern and Lausanne
According to the considerations reviewed above, the SLS have been used to produce a series of categorical maps for each urban agglomeration and time period.
The boundaries of each urban agglomeration have been adopted from the definitions provided by the Swiss Federal Statistical Office (SFSO) \cite{sfso2014espace}, which comprise multiple municipalities and have been established in consideration of population density, proximity between centers, economic activities and commuting behavior.
% The SLS inventory is managed within a relational database where each entry corresponds to the coordinates of a given pixel and its respective LULC class for each survey period\footref{fn:sls-years}. % which can be downloaded in a comma-separated value text file.
Since the SLS inventory is stored within a relational database where each entry corresponds to the coordinates of a given pixel and its respective land use/land cover classes for each survey period, an open source reusable toolbox to manage, transform and export categorical raster maps has been developed in Python \cite{bosch2019swisslandstats}.
% \cite{bosch2019swisslandstats}.
% The evolution of the urban agglomeration
The spatiotemporal evolution of the urban footprint for the three urban agglomerations is displayed in \autoref{figures/landscape_plots.pdf}.

\begin{figure}[!ht]
  \begin{adjustwidth}{-.4\textwidth}{0cm}
    \centering  
    \includegraphics[width=\linewidth]{figures/landscape_plots.pdf}
    % \vspace{.5em}
    \caption[Evolution of urban patches]{\label{figures/landscape_plots.pdf}Evolution of urban patches of the three urban agglomerations throughout their respective periods of study. The times t$_0$, t$_1$, t$_2$ and t$_3$ correspond to 1981, 1993, 2004 and 2013 for Bern; 1980, 1990, 2005 and 2014 for Lausanne and 1982, 1994, 2007 and 2016 for Zurich}
  \end{adjustwidth}
\end{figure}

\subsection*{Quantifying spatiotemporal patterns of urbanization}

% Several studies have undertaken the analysis of the spatio-temporal patterns of urbanization from the perspective of landscape ecology.
% TODO: (this in intro) Recent years have seen an increasing number of studies that employ landscape metrics to characterize the spatiotemporal patterns of LULC change associated to urbanization. % TODO: nameref to SI Table here
% While a plentiful collection of landscape metrics can be find in the literature, many of them are highly correlated with one another. As a matter of fact, \cite{riitters1995factor} found that the characteristics discerned by 55 prevalent landscape metrics could be reduced to only 6 independent factors. Accordingly, most of the reviewed studies employ a reduced number of metrics to characterize the spatial landscape patterns.
While a plentiful collection of landscape metrics can be find in the literature, many of them are highly correlated with one another. As a matter of fact, Riitters et al. \cite{riitters1995factor} found that the characteristics discerned by 55 prevalent landscape metrics could be reduced to only 6 independent factors.
On the other hand, landscape metrics can be very sensitive to the resolution and the extent of the maps. However, several metrics empirically exhibit consistent responses to changing scales that conform to predictable scaling relations \cite{wu2002empirical, wu2004effects}.
Based on such remarks, and in order to enhance comparability with other studies, ten landscape metrics have been selected for the present study, whose details are listed in \autoref{selected-metrics}.
While complying with the FRAGSTATS v4 definitions \cite{mcgarigal2012fragstats}, the landscape metrics have been computed with the open source library PyLandStats \cite{bosch2019pylandstats}.
Like in most of the related studies, the categorical maps have been reclassified into urban and natural classes, and the metrics have computed at the urban class level, namely aggregating their values across all the urban patches of the landscape. The only exceptions are the contagion (CONTAG) and Shannon's diversity index (SHDI), which might only be computed at the landscape level, namely considering all the landscape classes (urban and natural in the this study).
Pixels that correspond to land unavailable for development, such as water bodies, have been excluded from the computation of the metrics.

\begin{table}[!h]
  \begin{adjustwidth}{-.4\textwidth}{0cm} % Comment out/remove adjustwidth environment if table fits in text column.
    \footnotesize % Font size can be changed to match table content. Recommend 10 pt.
    \renewcommand{\arraystretch}{1.3} % horizontal spacing
    \centering
    \caption[Selected landscape metrics]{\label{selected-metrics}Selected landscape metrics to quantify the spatiotemporal patterns of urbanization. A more thorough description can be found in the documentation of the software FRAGSTATS v4 \cite{mcgarigal2012fragstats}}
    \begin{tabular}{p{.34\textwidth} p{.2\textwidth} p{.14\textwidth} p{.6\textwidth}} 
      \toprule
      \textbf{Metric name (abbrev.)} & \textbf{Level} & \textbf{Category} & \textbf{Description} \\
      \midrule
      Percentage of landscape (PLAND) & Class & Area and edge & Percentage of landscape, in terms of area, occupied by patches of a given class \\
      Mean patch size (MPS) & Class and landscape & Area and edge & Mean patch size \\
      Largest patch index (LPI) & Class and landscape & Area and edge & Percentage of the landscape occupied by the largest patch \\
      Patch density (PD) & Class and landscape & Aggregation & The number of patches per area unit \\
      Edge density (ED) & Class and landscape & Area and edge & Sum of the lengths of all edge segments per area unit \\
      Area-weighted mean fractal dimension (AWMFD) & Class and landscape & Shape & Mean patch fractal dimension weighted by relative patch area \\
      Mean euclidean nearest neighbor distance (ENN) & Class and landscape & Aggregation & Mean patch shortest edge-to-edge distance to the nearest neighboring patch of the same or different class \\
      Landscape shape index (LSI) & Class and landscape & Aggregation & Standardized ratio of edge length to area \\
      Contagion (CONTAG) & Landscape & Aggregation & Metric that measures the extent to which patches of the same class are spatially aggregated in the landscape \\
      Shannon's diversity index (SHDI) & Landscape & Diversity & Metric that measures the diversity of patch types determined by both the number of different patch types and the proportional distribution of area among patch classes. \\
      % SqP
      \bottomrule  
    \end{tabular}
  \end{adjustwidth}
\end{table}

\subsubsection*{Modes of urban growth}

% TODO: the two measures provide complementary information
In addition to the conventional landscape metrics, which are computed over a single snapshot of a landscape, Liu et al. \cite{liu2010new} proposed a quantitative method to classify the types of urban growth occurring between two time points. 
To that end, for each new urban patch, the Landscape Expansion Index (LEI) is computed as\footnote{The LEI definition of (\ref{eq:lei}) is taken from Nong et al. \cite{nong2018quantifying} and is equivalent to the initial formula proposed by Liu et al. \cite{liu2010new}}:

\begin{equation}
  \label{eq:lei}
  LEI = \frac{L_c}{P}
\end{equation}

where $L_c$ denotes the length of the interface between the new urban patch and pre-existing urban patches, and $P$ is the perimeter of the new urban patch. Then, the type urban growth attributed to a new urban patch will be identified as infilling when $LEI > 0.5$, edge-expansion when $0 < LEI \leq 0.5$ and leapfrog when $LEI = 0$.
% As suggested by \cite{li2013quantifying}, the relative dominance of each growth mode between two time points can be evaluated through the proportion of new urban patches that are attributed to each growth mode, both in terms of number and area of new urban patches.
As suggested by Li et al. \cite{li2013quantifying}, the relative dominance of each growth mode between two landscape snapshots can be evaluated both in terms of number and the area of new urban patches that are attributed to each growth mode.

\subsubsection*{Fractal aspects of urban patterns}

@@

% Fractal geometry has been extensively employed to describe morphological characteristics of cities and landscapes.
% Cities, like many other geographical phenomena, display complex morphological traits which despite their irregular appearance, comply with well-defined order principles that can be characterized quantitatively by means of fractal geometry.
% Cities, like many other geographical phenomena, display complex morphological traits.
Despite their complex and irregular appearance, cities comply with well-defined order principles that can be characterized quantitatively by means of fractal geometry \cite{frankhauser1994fractalite, batty1994fractal}.
% Two recurrent empirical observations are of particular interest in the context of the study of land use change associated to urbanization \cite{white2015modeling}.
Two main characteristics of fractal structures are of particular interest in the context of the study of urbanization and land use change \cite{white2015modeling}: the area-radius scaling and the size-frequency distribution of urban patches.

\paragraph*{Area-radius scaling}
% On the one hand, the radial dimension, which describes the relationship between the built-up area of an urban agglomeration and the distance from the main city center, has been shown to empirically conform to the relationship:
The relationship between the built-up area of an urban agglomeration and the distance from the main city center has been shown to empirically conform to the relationship:

@@

where $A$ denotes the total area of the urban built-up extent that lays within a distance $r$ from the city center, and $D$ corresponds to the radial dimension, analogous to the fractal dimension of two-dimensional complex objects such as Sierpinski carpets.
% Although the measure might not be appropriate for urban agglomerations with multiple important centers, \cite{frankhauser1994fractalite} found extensive evidence that not only most contemporary cities could be approximated through (\ref{eq:radial-dimension}), but also that with very few exceptions the value of $D$ fell between 1.9 and 2.0.
Although the measure might not be appropriate for urban agglomerations with multiple important centers, Frankhauser \cite{frankhauser1994fractalite} found extensive evidence that most contemporary cities could be approximated through (\ref{eq:radial-dimension}), with the value of $D$ consistently falling between 1.9 and 2.0.
% Additionally, as noted first by \cite{frankhauser1990aspects} and confirmed in subsequent examination by \cite{white1993cellular},
On the other hand, following initial observations by Frankhauser \cite{frankhauser1990aspects}, White and Engelen \cite{white1993cellular} suggested that the area-radius scaling could be better approximated throguh two values of $D$, a first steeper one for small values of $r$, reflecting an inner zone where urbanization was essentially complete, and a second lower slope for the outer zone that is still undergoing urbanization.

@@

\paragraph*{Size-frequency distribution of urban patches}
% On the other hand, the size distribution of urban patches
% On the other hand,
Contemporary urban agglomerations are configured by numerous patches of urban land use.
% For fractal objects, there must be no characteristic patch size
% A hallmark of fractal objects is the self-similarity accros scales, which means that there must be no characteristic patch size.
If such configuration is fractal, there must be no characteristic patch size, and thus the relationship between the size of an urban patch and the number of patches of that size found in the agglomeration must follow a power-law scaling relationship of the form:

\begin{equation}
  \label{eq:size-frequency}
  N(s) \sim s^{-\alpha}
\end{equation}

where $N(s)$ is the number of patches of size $s$, and the scaling exponent $\alpha$ is the patch size-frequency dimension \cite{white1993cellular}.

% TODO: such powerlaw scaling (f-noise, self-organized criticality [cite bak]) is characteristic of many social and natural systems [cite bettencourt]

@@

The relative dominance of infilling, edge expansion and leapfrog development during the period of study is displayed in \autoref{figures/growth_modes.pdf}.
% TODO: mention LEI and AWLEI here and in the figure caption
% Overall, considering the influence of growth modes in terms of the number of new urban patches rather than their respective area tends to over-represent the impact of leapfrog, since the new urban patches that appear non-adjacent to existing urban patches tend to be small.
% As suggested by \cite{li2013quantifying}, edge-expansion maintains its importance throughout the three periods considered for each agglomeration, whereas leapfrog and infilling alternate their relative dominance.
% As suggested by Li et al. \cite{li2013quantifying}, edge-expansion maintains its importance throughout all agglomerations and time periods, whereas leapfrog and infilling alternate their relative dominance.
% Nevertheless, the way in which the latter happens does not seem random in the present study but rather suggests a tendence towards increasing infill, more noticeable when considering its influence in terms of area.
% Nevertheless, in the present study, such alternation does not seem to occur randomly but rather suggests a tendence towards increasing infill, more noticeable when considering its influence in terms of area.
As suggested by Li et al. \cite{li2013quantifying}, all three urban growth modes act simultaneously.
Edge-expansion is the most dominant growth mode and maintains its importance throughout all agglomerations and time periods, nevertheless, it tends to diminish over time. Such a decrease of edge-expansion is compensated by an increasing influence of infilling. Leapfrog is by far the least dominant growth mode, and its influence does not show any clear trend of increase or decrease.
% A potential explanation might stem from the fact that suitable urban land is becoming incresingly scarce.
% See \nameref{code-sensitivity}.
